### 2021-06-08 paper/report.tex
Re-wrote most parts.

Since early 2020, the CoViD-19 pandemic has presented an enormous challenge to humanity
on many dimensions. The development of highly effective vaccines holds the promise of
containment in the medium term. However, most countries find themselves many
months---and often years---away from reaching vaccination-induced herd
immunity.\comment[id=HM]{Cite some paper on herd immunity, maybe vaccine data} In the
meantime, it is of utmost importance to employ an effective mix of strategies for
containing the virus. The most frequent initial response was a set of non-pharmaceutical
interventions (NPIs) to reduce contacts between individuals. While this has allowed some
countries to sustain equilibria with very low infection numbers\footnote{See
\citet{Contreras2021} for a theoretical equilibrium at low case numbers which is
sustained with test-trace-and-isolate policies.}, most have seen large ups and downs in
their infection rates. Containment measures have become increasingly diverse and
included testing, more nuanced NPIs, and contact tracing. Neither these policies' effect
nor the influence of seasonal patterns or more infectious virus strains are well
understood in quantitative terms. This paper develops a model incorporating all these
factors. The framework allows to combine a wide variety of data in a timely fashion,
making it useful to predict the effects of various interventions. We apply the model to
Germany and show that rapid testing had the largest impact on the reduction in
infections by almost 80\% during the month of May 2021. We conclude that rapid tests
have a large role to play at least as long as vaccinations have not been offered to an
entire population.

@@

@@

        \includegraphics[width=\textwidth]{../figures/model-graph-bottom-left}
        \caption{{\small Model for antigen tests}}
        \label{fig:antigen_tests}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.425\textwidth}
        \centering
        % \includegraphics[width=\textwidth]{../figures/model_detailed.png}
        Model for undetected cases
        \vskip4ex

        \caption{{\small Model for undetected cases}}
        \label{fig:model_for_official_cases}
    \end{subfigure}

@@

EQUATIONS FOR THE FOURTH PICTURE (find names!):

@@

\begin{align*}
    x &= \frac{n\_newly\_infected \times share\_known\_cases \times P(PCR | S)}{n\_symptomatic\_and\_untested} \\
    y &= \frac{n\_newly\_infected \times share\_known\_cases \times P(PCR | S)}{n\_asymptoptomatic\_and\_infected\_and\_untested} \\
    z &= \frac{P(PCR | RT)}{survey}
\end{align*}

where untested means that individuals are not waiting on the result of a previously
administered rapid test.

Susceptibility to contracting the SARS-CoV-2 virus is dependent on age; so is the
progression of a possible infection. We differentiate between an initial period of
infection without being infectious or showing symptoms, being infectious (presymptomatic
or asymptomatic), showing symptoms, requiring intensive care, and recovery or death. The
probabilities of transitioning between these states depend on age; their duration is
random within intervals calibrated to medical literature (for a detailed description see
Section~\ref{sec:medical_params}). Conditional on the type of contact, infectiousness
independent of ages \citep{Jones2021}.

The model includes several other features, which are crucial to describe the evolution
of the pandemic in 2020-2021. New virus strains with different profiles regarding
infectiousness \replaced[id=K]{}{and disease progress}\comment[id=K]{sid only supports
differing infectiousness between variants at the moment} can be introduced. With a
probability of 75\% \citep{Hunter2021}, vaccinated agents become immune and they do not
transmit the virus \citep{Petter2021, LevineTiefenbrun2021,
Pritchard2021}.\footnote{75\% is lower than what is usually reported for after the
second dose. We choose it because in our model does not include booster shots and does
not allow for vaccinated individuals to become infected and transmit the disease which
is the case for Covid-19 \citep{Petter2021, LevineTiefenbrun2021,
Pritchard2021}.}% the sources only report reduced viral shedding and reduced risk of infection.
% Lauterbach calculated that taking the risk of infection * reduced infectiousness given
% infection based on theses studies suggest that given exposure the risk that fully
% vaccinated individuals pose is <10% of an unvaccinated person.
While vaccines are rolled out, priority depends on age and occupation. Agents may demand
two types of tests as described in Figure~\ref{fig:antigen_tests}. First, polymerase
chain reaction (PCR) tests directly reveal whether an individual is infected or not. PCR
tests require some time to be processed and there are always aggregate capacity
constraints. Second, rapid antigen tests yield immediate results. Specificity and
sensitivity of these tests is set according to data by \cite{Bruemmer2021, Smith2021} .
The sensitivity of the rapid antigen tests depends on the timing of the test relative to
the start of infectiousness.\comment[id=HM]{subsequent PCR tests, quarantine,
work/school}. Demand for rapid tests will depend on policy, e.g., prices, testing
requirements before work, attending school, or visiting restaurants and shops.

Modelling a population of agents according to actual demographic characteristics means
that we can use a wide array of data to identify and estimate the model's many
parameters.\footnote{See section S.X of the supplementary materials for an
overview.}\comment[id=HM]{Need to make a list including symbols, maybe you can fetch the
parameters from the model and then we brainstorm re notation?} Mobility data is used to
model the reductions in work contacts. School and daycare policies are incorporated
directly from official directives. Infection rates by age and geographical region are
estimated to match to officially recorded numbers; so is the prevalence of virus
strains. In order to translate simulated cases in our model to officially recorded
cases, we use the model depicted in
Figure~\ref{fig:model_for_official_cases}.\comment[id=HM]{Add a sentence.} A further
advantage is that the simulated data have a structure that resembles datasets used for
regression models, which allows additional plausibility checks by re-running the same
models on the model-generated data.\comment[id=HM]{Only keep if we actually do something
like that.}

Model that makes the most of many available data sources to gauge the relative effects
in this transition period -- see joint distribution of infections with age, geography.
But contacts with age, geography, occupations.

We apply this model to Germany. In March and April 2020, the country broke the first
wave of the pandemic fairly quickly. Between mid-May and mid-September, daily new
infections were below 20 per Million and day.\comment[id=HM]{Cite our world in data} We
model the period mid-September 2020 to the end of May 2020. We pick the starting date
for two reasons. First, we do not include the first wave because the environment was
very different (e.g., aggregate PCR test capacity was much lower and we would require a
very different model for calculating the share of known cases) and some data was not
recorded yet.\comment[id=HM]{True?}\comment[id=K]{We would definitely need to model
national and international travel which you mention below. Regarding data I think
everything would be available. The low numbers during summer in our sample size might
also be a problem because Covid might go extinct in our model when case numbers are so
low.}. Second, a large fraction cases during summer of 2020 were traced to international
travel \citep{KochInstitut2021}, but the precise number is difficult to model.

Figure~\ref{fig:pandemic_drivers_model_fit} describes the evolution of the pandemic and
of its drivers. The black line in Figure~\ref{fig:aggregated_fit} shows officially
recorded cases; the \replaced[id=HM]{color}{blue?} line in
Figure~\ref{fig:stringency_index} the Oxford Response Stringency Index \citep{Hale2020},
which tracks the tightness of non-pharmaceutical interventions. We transform the index
so that lower values represent higher levels of restrictions. A value of zero means all
measures incorporated in the index are turned on. The value 1 represents the situation
of mid-September, where half of all restrictions incorporated in the index are
active.\comment[id=HM]{Need to find version of index with a couple of measures, probably
need the API access...}. In the six weeks between mid September and late October, cases
increased by a factor of
4.% the week of Sep 13 had an incidence of 12.4, the week of Oct 18 had 51.
Restrictions were somewhat tightened in mid-October and again in early November. New
infections remained constant throughout November, before rising again in December.

@@

Quickly rising rates in December prompted the most stringent lockdown to this date.
Schools and daycare centers were closed again, so were customer-facing businesses except
for grocery and drug stores. Shortly afterwards, more factors become relevant. Just
after Christmas, the first people were vaccinated with a focus on older age groups and
medical personnel (Figure~\ref{fig:antigen_tests_vaccinations}. By the end of May, just
over 40\% had received at least one dose of a vaccine. Starting in January, rapid tests
\dots These had to be administered by medical doctors or in pharmacies. At-home tests
approved by authorities were ... in March.\comment[id=HM]{update precise
timeline}\comment[id=K]{From a docstring:Bürgertests started in mid March but demand was
very low initially (\url{https://bit.ly/3ehmGcj}). First tests to self-administer became
available starting March 6. However, supply was very limited in the beginning
(\url{https://bit.ly/3xJCIn8}). The first tests were licensed on Feb 24:
\url{https://www.bundesregierung.de/breg-de/themen/coronavirus/zulassung-schnell-test-1861354}}
From the peak of the second wave just before Christmas until the trough in mid-February,
newly detected cases decreased by almost three quarters. The third wave in the spring of
2021 is associated with a rise in the share of the B.1.1.7 variant among all infections,
depicted in Figure~\ref{fig:share_b117}, reaching more than 80\% in
April\comment[id=HM]{Do we have the table somewhere?}\comment[id=K]{is in the empirical
analogues table}. In early March, some NPIs were being relaxed; e.g., hairdressers and
home improvement stores were allowed to open again to the public. In some cases,
customers were only allowed to enter with a recent negative rapid test result.

These developments are characteristic of many countries: The initial focus on NPIs to
slow the spread of the disease has been accompanied by vaccines and a growing acceptance
and use of rapid tests. At broadly similar points in time, novel strains of the virus
have started to pose additional challenges. The blue line in
Figure~\ref{fig:aggregated_fit} shows that our model tracks the simulated infections
extremely well. This is also true for infections by age and geographical region, which
are shown in the supplementary materials (figures~\ref{fig:age_group_fit,
fig:state_fit}. We can disentangle various mechanisms due to the distinct temporal
variation in the drivers of the pandemic. Next to the stringency index, the three lines
in Figure~\ref{fig:stringency_index} summarize how contact reductions, increased hygiene
regulations, and seasonality evolved since early September for each of the three broad
contact networks. For example, a value of 0.75 for the work multiplier means that if the
environment was the same as in September (levels of infection rates, no rapid tests or
vaccinations, only the wildtype virus present), infections at the workplace would be
reduced by 25\%. This reduction is the product of the effect of contact reductions,
increased hygiene regulations, and seasonality. Along with the levels of infections,
these measures completely determine the spread of SARS-CoV-2 in
2020.\comment[id=HM]{True?} See Appendix~\comment[id=HM]{Reference} \comment[id=K]{We
have not done the decomposition for 2020 yet. Is on my todo list} for their individual
contributions. Two things are particularly interesting. First, all lines broadly follow
the stringency index and they would do so even more if we left out seasonality. Second,
the most stringent regulations are associated with the period of strong decreases in new
infection between late December 2020 and mid-February 2021. The measures were not
enough, however, to stop the B.1.1.7 variant from spreading in the subsequent period.
The steep drop in recorded cases during May 2021 is associated with rapid tests to
X-Y~percent of the population, a vaccination rate that rose from 25\% to 4X\%, and
mostly seasonality impacting a fall in the relative infectiousness of contacts outside
of work and school\comment[id=HM]{check with corrected graph}.

@@

\paragraph{Points to mention}
\begin{itemize}
    \item If anything too optimistic regarding vaccinations
    \item Social structure / conditional block testing in families important (?)
    \item Trump-effect: More testing = more cases true for how long?
\end{itemize}

>>>>

Since early 2020, the CoViD-19 pandemic has presented an enormous challenge to humanity
on many dimensions. The development of highly effective vaccines holds the promise of
containment in the medium term. However, most countries find themselves many
months---and often years---away from reaching vaccination-induced herd
immunity.\comment[id=HM]{Cite some paper on herd immunity, maybe vaccine data} In the
meantime, it is of utmost importance to employ an effective mix of strategies for
containing the virus. The most frequent initial response was a set of non-pharmaceutical
interventions (NPIs) to reduce contacts between individuals. While this has allowed some
countries to sustain equilibria with very low infection numbers\footnote{See
\citet{Contreras2021} for a theoretical equilibrium at low case numbers which is
sustained with test-trace-and-isolate policies.}, most have seen large fluctuations of
infection rates. Containment measures have become increasingly diverse and now include
testing, more nuanced NPIs, and contact tracing. Neither these policies' effect nor the
influence of seasonal patterns or more infectious virus strains are well understood in
quantitative terms. This paper develops a model incorporating all these factors. The
framework allows to combine a wide variety of data in a timely fashion, making it useful
to predict the effects of various interventions. We apply the model to Germany and show
that rapid testing had the largest impact on the reduction in infections by almost 80\%
during the month of May 2021. We conclude that rapid tests have a large role to play at
least as long as vaccinations have not been offered to an entire population.

@@

\begin{figure}
    \centering

@@

        \includegraphics[width=\textwidth]{../figures/model-graph-bottom-left}
        \caption{{\small Model for antigen tests}}
        \label{fig:antigen_tests}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.425\textwidth}
        \centering
        \includegraphics[width=\textwidth]{../figures/model-graph-bottom-right}
        \caption{{\small Model for undetected cases}}
        \label{fig:model_for_official_cases}
    \end{subfigure}

@@

In our model, susceptibility to contracting the SARS-CoV-2 virus is dependent on age. A
possible infection progresses as shown in Figure~\ref{fig:disease_progression}. We
differentiate between an initial period of infection without being infectious or showing
symptoms, being infectious (presymptomatic or asymptomatic), showing symptoms, requiring
intensive care, and recovery or death. The probabilities of transitioning between these
states depend on age; their duration is random within intervals calibrated to medical
literature (for a detailed description see Section~\ref{sec:medical_params}).
Conditional on the type of contact, infectiousness independent of ages
\citep{Jones2021}.\comment[id=HM]{Seems broadly standard, can we cite someone?}

The model includes several other features, which are crucial to describe the evolution
of the pandemic in 2020-2021. New virus strains with different profiles regarding
infectiousness can be introduced. Agents may receive a vaccination. With a probability
of 75\% \citep{Hunter2021}, vaccinated agents become immune and they do not transmit the
virus \citep{Petter2021, LevineTiefenbrun2021, Pritchard2021}.\footnote{75\% is lower
than what is usually reported for after the second dose of the Biontech/Pfizer vaccine,
which is most commonly used in Germany. We choose it because our model neither includes
booster shots, nor does it allow vaccinated individuals who became immune to transmit
the disease\citep{Petter2021, LevineTiefenbrun2021, Pritchard2021}.} During the vaccine
roll-out, priority may depend on age and occupation.

We include two types of tests. Polymerase chain reaction (PCR) tests directly reveal
whether an individual is infected or not. PCR tests require some days to be processed
and there are always aggregate capacity constraints. In contrast, rapid antigen tests
yield immediate results. Specificity and sensitivity of these tests is set according to
data analysed in \cite{Bruemmer2021, Smith2021}; sensitivity depends on the timing of
the test relative to the start of infectiousness. Figure~\ref{fig:antigen_tests} shows
our model for rapid test demand. Schools may require students to be tested regularly.
Rapid tests may be offered by employers for on-site workers. Individuals may demand
tests for private reasons, which include having plans to meet other people\footnote{A
positive test will make them reduce their contacts; this is why tests impact the actual
contacts in Figure~\ref{fig:model-description}.}, showing symptoms of CoViD-19, and
because a household member tested positively for the virus. We endow each agent with an
individual compliance parameter. This parameter determines whether she takes up rapid
tests offered by employers or follows up on private reasons. Thresholds differ across
reasons triggering rapid test demand; they are lowest for infected household members and
highest for meeting other people.\comment[id=HM]{True?}.

Modelling a population of agents according to actual demographic characteristics means
that we can use a wide array of data to identify and estimate the model's many
parameters.\footnote{See section S.X of the supplementary materials for an
overview.}\comment[id=HM]{Need to make a list including symbols, maybe you can fetch the
parameters from the model and then we brainstorm re notation?} Mobility data is used to
model the reductions in work contacts. School and daycare policies are incorporated
directly from official directives. Infection rates by age and geographical region are
estimated to match to officially recorded numbers; so is the prevalence of virus
strains. The model yields total infections, but only a fraction of those will be
officially recorded. We model the fraction of known cases as depicted in
Figure~\ref{fig:model_for_official_cases}.\comment[id=HM]{Add a sentence.}
\footnote{EQUATIONS FOR THE FOURTH PICTURE (find names!):

@@

    \begin{align*}
        x & = \frac{n\_newly\_infected \times share\_known\_cases \times P(PCR | S)}{n\_symptomatic\_and\_untested}                    \\
        y & = \frac{n\_newly\_infected \times share\_known\_cases \times P(PCR | S)}{n\_asymptoptomatic\_and\_infected\_and\_untested} \\
        z & = \frac{P(PCR | RT)}{survey}
    \end{align*}
    where untested means that individuals are not waiting on the result of a previously
    administered rapid test.} A further advantage is that the simulated data have a
    structure that resembles datasets used for regression models, which allows
    additional plausibility checks by re-running the same models on the model-generated
    data.\comment[id=HM]{Only keep if we actually do something like that.}

We apply this model to Germany. In March and April 2020, the country broke the first
wave of the pandemic fairly quickly. Between mid-May and mid-September, daily new
infections were below 20 per Million and day.\comment[id=HM]{Cite our world in data} We
model the period mid-September 2020 to the end of May 2021. We pick the starting date
for two reasons. First, we do not include the first wave because the environment was
very different (e.g., aggregate PCR test capacity was much lower and we would require a
very different model for calculating the share of known cases) Second, a large fraction
cases during summer of 2020 were traced to international travel
\citep{KochInstitut2021}\comment[id=HM]{Include
\url{https://www.nature.com/articles/s41586-021-03677-y} in citation}, but the precise
number is difficult to model.

Figure~\ref{fig:pandemic_drivers_model_fit} describes the evolution of the pandemic and
of its drivers. The black line in Figure~\ref{fig:aggregated_fit} shows officially
recorded cases; the \replaced[id=HM]{color}{blue?} line in
Figure~\ref{fig:stringency_index} the Oxford Response Stringency Index \citep{Hale2020},
which tracks the tightness of non-pharmaceutical interventions. We transform the index
so that lower values represent higher levels of restrictions. A value of zero means all
measures incorporated in the index are turned on. The value 1 represents the situation
in mid-September, with restrictions on gatherings and public events, masking
requirements, but open schools and workplaces (the raw value of the index at that point
is 49.5). In the seven weeks between mid September and early November, cases increased
by a factor of 10. Restrictions were somewhat tightened in mid-October and again in
early November. New infections remained constant throughout November, before rising
again in December, which prompted the most stringent lockdown to this date. Schools and
daycare centers were closed again, so were customer-facing businesses except for grocery
and drug stores. From the peak of the second wave just before Christmas until the trough
in mid-February, newly detected cases decreased by almost three quarters. The third wave
in the spring of 2021 is associated with the B.1.1.7 strain, which became dominant in
March. See Figure~\ref{fig:share_b117}. In early March, some NPIs were being relaxed;
e.g., hairdressers and home improvement stores were allowed to open again to the public.

@@

By this time, the set of policy instruments had become much more diverse. Around the
turn of the year, the first people were vaccinated with a focus on older age groups and
medical personnel (Figure~\ref{fig:antigen_tests_vaccinations}. By the end of May, just
over 40\% had received at least one dose of a vaccine. Around the same time, rapid tests
started to replace PCR tests in many medical and nursing facilities. These had to be
administered by medical doctors or in pharmacies. At-home tests approved by authorities
became available in mid-March, rapid test centers were opened and one test per person
and week was made available free of charge. Depending on the state, customers were only
allowed to enter certain stores with a recent negative rapid test result. These
developments are characteristic of many countries: The initial focus on NPIs to slow the
spread of the disease has been accompanied by vaccines and a growing acceptance and use
of rapid tests. At broadly similar points in time, novel strains of the virus have
started to pose additional challenges.

Our model is able to track these developments very well. The blue line in
Figure~\ref{fig:aggregated_fit} shows our model's predictions are very close to
officially recorded cases in the aggregate. This is also true for infections by age and
geographical region, which are shown in the supplementary materials
(Figures~\ref{fig:age_group_fit} and \ref{fig:state_fit}, respectively). We can
disentangle various mechanisms due to the distinct temporal variation in the drivers of
the pandemic. Next to the stringency index, the three lines in
Figure~\ref{fig:stringency_index} summarize how contact reductions, increased hygiene
regulations, and seasonality evolved since early September for each of the three broad
contact networks. For example, a value of 0.75 for the work multiplier means that if the
environment was the same as in September (levels of infection rates, no rapid tests or
vaccinations, only the wildtype virus present), infections at the workplace would be
reduced by 25\%. This reduction is the product of the effect of contact reductions,
increased hygiene regulations, and seasonality. Along with the levels of infections,
these measures completely determine the spread of SARS-CoV-2 in
2020.\comment[id=HM]{True?} See Appendix~\comment[id=HM]{Reference}
\comment[id=K]{We have not done the decomposition for 2020 yet. Is on my todo list}
\comment[id=HM]{I guess my question was whether I was missing any factors.}
for their individual contributions. Two things are particularly interesting. First, all
lines broadly follow the stringency index and they would do so even more if we left out
seasonality and school vacations (roughly the last two weeks of October, two weeks each
around Christmas and Easter, and some days in late May). Second, the most stringent
regulations are associated with the period of strong decreases in new infection between
late December 2020 and mid-February 2021. The measures were not enough, however, to stop
the B.1.1.7 variant from spreading in the subsequent period. The steep drop in recorded
cases during May 2021 is associated with at least weekly rapid tests to around
42~percent of the population, a vaccination rate that rose from 28\% to 43\%, and mostly
seasonality impacting a fall in the relative infectiousness of contacts outside of work
and school.

@@

\paragraph{Points to mention}
\begin{itemize}
    \item If anything too optimistic regarding vaccinations
    \item Social structure / conditional block testing in families important (?)
    \item Trump-effect: More testing = more cases true for how long?
    \item Towards the end, total tests go down: Preferences for tests remain the same,
    but infection rates lower \comment[id=HM]{Janos, please fill up}
\end{itemize}
